The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-07-17 WavesData-2019/LatexNotes/notes.tex
NM

% Basic physical parameters and scales
\newcommand{\freqp}{f_p}
\newcommand{\amp}{a}
%\newcommand{\etastd}{\eta_{\text std}}
% Parameters that change crossing the ADC
\newcommand{\depth}{d}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\dupdn}{\depth_{\pm}}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamdn}{\lam_{+}}
\newcommand{\lamupdn}{\lam_{\pm}}
\newcommand{\lamfac}{N}
\newcommand{\drat}{D}
\newcommand{\dratdn}{\drat_+}
\newcommand{\dratupdn}{\drat_{\pm}}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
\newcommand{\En}{\mathcal{E}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}

@@

We introduce the following characteristic scales
\begin{align}
\label{ampdefn}
&\amp = \mean{\eta^2}^{1/2} 
&&\mbox{\em characteristic wave amplitude} \\
\label{lamdefn}
&\lamupdn = \sqrt{g \dupdn} / \freqp
&&\mbox{\em characteristic wave length}
\end{align}
where $\mean{\cdot}$ indicates the mean of a quantity. Note that the characteristic wavelength $\lam = \lamupdn$ takes different values upstream and downstream of the ADC. We remark that experimental measurements indicate that the characteristic amplitude $\amp = \mean{\eta^2}^{1/2}$ is nearly the same on both values of the ADC. Hence, we will not distinguish between upstream and downstream values of $\amp$.
We introduce the following dimensionless variables
\begin{align}
&u = \eta/\amp
&&\mbox{\em dimensionless surface displacement} \\
&\tilde{x} = \frac{\pi \xi}{\lamfac \lam} = \frac{\pi (x-ct)}{\lamfac \lam}
&&\mbox{\em dimensionless position in moving frame} \\
& \tilde{t} = t \freqp / \lamfac
&&\mbox{\em dimensionless time}
\end{align}
We will consider the physical domain $\xi \in [-\lamfac \lam, \lamfac \lam]$, which is an integer multiple of the characteristic wavelength, where $\lamfac$ will be selected later (probably about 5). Above, we chosen to normalize position by the scale $L = \lamfac \lam/\pi$ so that the dimensionless domain is $\tilde{x} \in [-\pi,\pi]$. Accordingly, we have also chosen the time scale as $\lamfac/\freqp$, so that a wave traveling at the local wavespeed $c = \lam \freqp = \sqrt{g \depth}$ crosses the physical domain of length $2 \lamfac \lam$ in one unit of dimensionless time. Said differently, the dimensionless wave speed is $(2 \pi)^{-1}$ so that a wave crosses the domain $[-\pi,\pi]$ in unit time.

@@

Dropping the tildes, the dimensionless KdV equation takes the form
\begin{equation}
\label{dimlessKdV}
u_t + {3 \pi} \epsup \drat^{-1} \, u u_x + \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where dimensionless variables are as follows
\begin{align}
&\epsup = \amp/\dup \, , \\
&\delup = \dup/\lamup = \sqrt{\dup \freqp^2/g} \, \\
&\drat = {\depth}/{\dup} \, .
\end{align}
Here, $\drat = \dratupdn$ is the local dimensionless depth, which changes value crossing the ADC. In particular, since \eqref{dimlessKdV} is in the reference frame that moves with the characteristic wave speed, we can assume that the ADC is met at $t = T_{ADC} = 0$. Then we have
\begin{equation}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } t<0 \\
\dratdn 	&\quad \mbox{for } t>0
\end{cases}
\end{equation}
where $\dratdn = \ddn/\dup$.
See Table \ref{paramtable} for a summary of important parameters.

@@

To compare to the previous formulation, we note that
\begin{align}
\amp \equiv \mean{\eta^2}^{1/2} = \sqrt{ \frac{E}{\pi L} }
\end{align}
So that the new normalization factor is
\begin{equation}
E^{1/2}L^{-1/2} = \sqrt{\pi} \amp
\end{equation}
In other words, the normalizations for surface displacement used here versus that in Section \ref{nondim}
 differ by a factor of $\sqrt{\pi}$. Fortunately, the normalization for position is identical in the two. We do have to remember that the normalization factor for position $L = \lamfac \lam / \pi$ takes two different values on either side of the ADC.

TO DO: How to scale time ....

@@

We now consider the Hamiltonian structure of the dimensionless variable-depth KdV equation \eqref{dimlessKdV}, over dimensionless domain $x \in [-\pi,\pi]$. First, the spatial average of any quantity is given by
\begin{equation}
\mean{q} = \frac{1}{2\pi} \int_{-\pi}^{\pi} q \dx \, .
\end{equation}
Appealing to near Ergodicity, we will typically assume equivalence between spatial, temporal, and ensemble averages.

@@

Then the Hamiltonian of \eqref{dimlessKdV} is given by
\begin{equation}
\Ham = 3 \pi \epsup \drat^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, H_2
\end{equation}
where $\drat$ changes value across the ADC. More specifically, there exists an upstream and a downstream Hamiltonian
\begin{align}
&\Hup = 3 \pi \epsup \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \, H_2 \\
&\Hdn = 3 \pi \epsup \dratdn^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \dratdn \, H_2
\end{align}
$\Hup$ is a conserved quantity for $t<0$ and $\Hdn$ is a conserved quantity for $t>0$. Each Hamiltonian $\Ham^{\pm}$ generates a corresponding Gibbs measure $\Gibbs^{\pm}$ with corresponding ensemble means $\mean{}_{\pm}$. (FILL IN DEFINITIONS)

>>>>

% Basic physical parameters and scales
\newcommand{\freqp}{f_p}
\newcommand{\amp}{a}
\newcommand{\ampscale}{A}
%\newcommand{\etastd}{\eta_{\text std}}
% Parameters that change crossing the ADC
\newcommand{\depth}{d}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\dupdn}{\depth_{\pm}}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamdn}{\lam_{+}}
\newcommand{\lamupdn}{\lam_{\pm}}
\newcommand{\lamfac}{N}
\newcommand{\drat}{D}
\newcommand{\dratdn}{\drat_+}
\newcommand{\dratupdn}{\drat_{\pm}}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
\newcommand{\En}{\mathcal{E}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}

@@

We introduce the following characteristic scales
\begin{align}
\label{ampdefn}
&\amp = \mean{\eta^2}^{1/2} 
&&\mbox{\em characteristic wave amplitude} \\
\label{lamdefn}
&\lamupdn = \sqrt{g \dupdn} / \freqp
&&\mbox{\em characteristic wave length}
\end{align}
where $\mean{\cdot}$ indicates the mean of a quantity. Note that the characteristic wavelength $\lam = \lamupdn$ takes different values upstream and downstream of the ADC. We remark that experimental measurements indicate that the characteristic amplitude $\amp = \mean{\eta^2}^{1/2}$ is nearly the same on both values of the ADC. Hence, we will not distinguish between upstream and downstream values of $\amp$.
We introduce the following dimensionless variables
\begin{align}
&u = \eta/\amp
&&\mbox{\em dimensionless surface displacement} \\
&\tilde{x} = \frac{\pi \xi}{\lamfac \lam} = \frac{\pi (x-ct)}{\lamfac \lam}
&&\mbox{\em dimensionless position in moving frame} \\
& \tilde{t} = t \freqp / \lamfac
&&\mbox{\em dimensionless time}
\end{align}
We will consider the physical domain $\xi \in [-\lamfac \lam, \lamfac \lam]$, which is an integer multiple of the characteristic wavelength, where $\lamfac$ will be selected later (probably about 5). Above, we chosen to normalize position by the scale $L = \lamfac \lam/\pi$ so that the dimensionless domain is $\tilde{x} \in [-\pi,\pi]$.

Accordingly, we have also chosen the time scale as $\lamfac/\freqp$, so that a wave traveling at the local wavespeed $c = \lam \freqp = \sqrt{g \depth}$ crosses the physical domain of length $2 \lamfac \lam$ in one unit of dimensionless time. Said differently, the dimensionless wave speed is $(2 \pi)^{-1}$ so that a wave crosses the domain $[-\pi,\pi]$ in unit time. Note: perhaps it is reasonable to normalize time differently, so that it is 'slow' time. Looking in Johnson, there does not seem to be a standard way to do this, it is very haphazard. So lets leave it for now.

@@

Dropping the tildes, the dimensionless KdV equation takes the form
\begin{equation}
\label{dimlessKdV1}
u_t + \frac{3}{2} \pi \epsup \drat^{-1} \, u u_x + \frac{1}{6} \pi^3 \lamfac^{-2} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where dimensionless variables are as follows
\begin{align}
&\epsup = \amp/\dup \, , \\
&\delup = \dup/\lamup = \sqrt{\dup \freqp^2/g} \, \\
&\drat = {\depth}/{\dup} \, .
\end{align}
Here, $\drat = \dratupdn$ is the local dimensionless depth, which changes value crossing the ADC. In particular, since \eqref{dimlessKdV1} is in the reference frame that moves with the characteristic wave speed, we can assume that the ADC is met at $t = T_{ADC} = 0$. Then we have
\begin{equation}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } t<0 \\
\dratdn 	&\quad \mbox{for } t>0
\end{cases}
\end{equation}
where $\dratdn = \ddn/\dup$.
See Table \ref{paramtable} for a summary of important parameters.

@@

To compare to the previous formulation, we note that
\begin{align}
\amp \equiv \mean{\eta^2}^{1/2} = \sqrt{ \frac{E}{\pi L} }
\end{align}
So that the new normalization factor for $\eta$ is
\begin{equation}
E^{1/2}L^{-1/2} = \sqrt{\pi} \amp
\end{equation}
In other words, the normalizations for surface displacement used here versus that in Section \ref{nondim}
 differ by a factor of $\sqrt{\pi}$. Fortunately, the normalization for position is identical in the two. We do have to remember that the normalization factor for position $L = \lamfac \lam / \pi$ takes two different values on either side of the ADC.

Lets do the nondimensionalization generally, so that we can easily make modifications to what scales to use. In general, we introduce dimensionless variables as
\begin{align}
&u = \eta / \ampscale \\
&\tilde{x} = x / L \\
&\tilde{t} = t / T
\end{align}
where $\ampscale, L, T$ are the scales for surface displacement (amplitude), length, and time respectively, which we are free to choose later. Then the dimensionless KdV equation is
\begin{equation}
u_t + \frac{3}{2} \left( \frac{c T \ampscale}{\depth L} \right) u u_x 
+ \frac{1}{6} \left( \frac{c \depth^2 T}{L^3} \right) u_{xxx} = 0
\end{equation}
where $c = \sqrt{g \depth}$.
In particular, choosing $\ampscale = \amp$, $L = \lamfac \lam/\pi$, and $T = \lamfac/\freqp$ produces \eqref{dimlessKdV1}. To see this, one has to recall $\lam = c/\freqp = \sqrt{g \depth}/\freqp$, and also has to collect all of the factors of $\depth$ that appear directly and appear in $L$.

Alternatively, we would like to choose $\ampscale = E^{1/2} L^{-1/2} = \sqrt{\pi} \amp$, $L = \lamfac \lam/ \pi$, and not sure yet about $T$. Ahh, I tried, but I cannot make sense of this because I cannot discern if $L$ in the PNAS paper and J Stat Phys paper is supposed to be dimensional or dimensionless. Also, I cannot tell of $L$ is supposed to change value crossing the ADC. The wavelength $\lam$ certainly changes value across the ADC, which suggests that $L$ should too, but on the other hand $L$ is treated as a constant throughout in the PNAS paper. So is $L$ the upstream value? I am not sure.

% Possible Flaw?
%Note: I believe this normalization is flawed because, for periodicity of $u$ to be maintained, the scale $L$ must change across the ADC. In particular, $L \sim \sqrt{\depth}$. However, the equations in our PNAS paper treat $L$ as constant throughout, even when $\drat$ changes. This seems to be a flaw to me.
% Maybe not a flow if you remember that $E^{1/2} L^{-1/2} = \sqrt{pi} \amp$ is the same on either side. Thus as L changes because wavelength changes, E changes accordingly.

Therefore, lets stick with the original normalization \eqref{dimlessKdV1}, repeated here
\begin{equation}
\label{dimlessKdV2}
u_t + \frac{3}{2} \pi \epsup \drat^{-1} \, u u_x + \frac{1}{6} \pi^3 \lamfac^{-2} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
If we need to draw a comparison to the PNAS paper, we set something like $\epsup = E^{1/2}L^{-3/2}$ and $\delup^2 = L^{-3}$, etc.

@@

We now consider the Hamiltonian structure of the dimensionless variable-depth KdV equation \eqref{dimlessKdV1}, over dimensionless domain $x \in [-\pi,\pi]$. First, the spatial average of any quantity is given by
\begin{equation}
\mean{q} = \frac{1}{2\pi} \int_{-\pi}^{\pi} q \dx \, .
\end{equation}
Appealing to near Ergodicity, we will typically assume equivalence between spatial, temporal, and ensemble averages.

@@

Then the Hamiltonian of \eqref{dimlessKdV1} is given by
\begin{equation}
\Ham = 3 \pi \epsup \drat^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, H_2
\end{equation}
where $\drat$ changes value across the ADC. More specifically, there exists an upstream and a downstream Hamiltonian
\begin{align}
&\Hup = 3 \pi \epsup \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \, H_2 \\
&\Hdn = 3 \pi \epsup \dratdn^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \dratdn \, H_2
\end{align}
$\Hup$ is a conserved quantity for $t<0$ and $\Hdn$ is a conserved quantity for $t>0$. Each Hamiltonian $\Ham^{\pm}$ generates a corresponding Gibbs measure $\Gibbs^{\pm}$ with corresponding ensemble means $\mean{}_{\pm}$.

(FILL IN DEFINITIONS)
